Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-05 paper_jasa.tex
typo fixed

After many small pilot studies with local personnel, it was clear that some care was needed to set up the human subjects experiments. It was best for a subject to see a block of 10 lineups with varying difficulty, with a reasonable number of ``easy'' lineups. The explanations about each experiment (below) includes an explanation of how the lineups were sampled and provided to the observers, for each of the three experiments.

@@

The experiment is designed to study the ability of human observers to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated a range of values of $\beta_2~ (=0, 1, 3, 5, 7, 8, 10, 16)$, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical conventional test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making a side-by-side boxplot of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and these were plotted in the same way as side-by-side boxplots. The actual data plot was randomly placed among these null data plots to produce the lineup. Figure \ref{fig:test_category} is an example of one of these lineups. The actual data plot is number 11 which was generated for $n$=300, $\beta_2$=15 and $\sigma$=12.

@@

%With multiple observers it is possible to estimate the visual $p$-value, and a measure of plot signal strength. The visual $p$-value is calculated from the binomial formula (Equation \ref{binom}). Assuming $H_o$ is true, the probability of picking the actual data plot by chance is $1/m$, and the visual $p$-value is the probability that $x$ or more observers select that plot. Figure \ref{fig:pval_definition} displays this visual $p$-value for each lineup against the conventional $p$value.

%Similary, the strength of the signal, is calculated using the proportion of observers picking actual data plot. It is closely related to the power of the test. From Equation \ref{eqn:power_estimate}  we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives the signal strength estimate as  
%\[
%\hat p_D= (1- {u}/{K})/({m-1})
%\]
%where $K$ is the total number of independent evaluations on a lineup and $u$ is the number of successful evaluations. Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with conventional test $p$-values, on a square root scale. For experiment 3, technically, there are no conventional $p$-values, but the $p$-values are computed in the same manner as experiment 2, assuming that the data satifies the assumptions. The blue line is a loess fit to the data, and the grey line indicates equality of the two axes. The relationship between the conventional $p$-value and the plot signal strength is very strong, particularly when the values are small. As the conventional $p$-value increases, the estimated signal strength increases -- observers find it harder to detect the actual data plot at larger $p$-values. The rise is very sharp for experiment 1, which indicates that observers have a harder time detecting the actual data plot for even small $p$-values.

@@

One assumption made in order to calculate the effect of lineup size in the calculations of visual $p$-value and signal strength was that subjects would tend to pick the plot in the lineup that had the strongest signal. In experiments 1 and 2, this corresponds to the plot with the smallest $p$-value. This can be examined for the data collected, to see if this assumption is, indeed, reasonable. %In Section \ref{sec:size} in order to make comparisons with conventional tests an assumption was made that the observer would pick the plot in the lineup that had the lowest $p$-value. In this section we take a look at the picks that subjects made, to see if this is, indeed, true.

Figures \ref{fig:P-val_log} and \ref{fig:P-val_log2} show the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizontally and the count of the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. Each row corresponds to a different $\beta$ value ordered top to bottom from lowest to highest. Columns correspond to the other levels of the experiment, sample size $(n)$, error standard deviation $(\sigma)$ and replicate. (Empty cells indicate no lineup shown with this combination.) Red indicates the plot with the lowest $p$-value in the lineup. In both experiments as $\beta$ gets larger (stronger effect) people tended to select the plot with the lowest $p$-value. This is less so the case when $\beta$ is small, when standard deviation is large and sample size is small. The results are clearer for experiment 2, that used a continuous covariate. But even when they didn't pick the plot with the lowest $p$-value they tended to oscillate their choices between the several low $p$-value plots. So for most observers, the assumption that they pick the plot with the smallest $p$-value would appear to be reasonable, and the actual power of the visual test should be close to the expected power.

@@

There are some noticeable exceptions to this rule. In experiment 1, when $\beta=5, n=300, sd=12, rep=3$, people tended to pick the plot with the second smallest $p$-value.  Same experiment, when $\beta=0, n=100, sd=5, rep=1$ people overwhelming chose a plot with much larger $p$-value. For several of these exceptions, along with several easy lineups, we followed up with a new experiment using an eye-tracker to examine what the observers were looking at in making their choices. \cite{*** yifan's CC needs to be made into a tech report, and referenced here.}

Experiment 3 does not have $p$-values associated with each plot, so the assumption can only be tested with the first two experiments. Practically, this is not an assumption that is explicitly made, because there will not be a $p$-value available. Rather it is assumed that the observer has the ability to select the plot with the most extreme structure. A few of the lineups indicated that occasionally people cue on different patterns, for example, in experiment 1 the lineup $\beta=0, n=100, sd=5, rep=1$ (first row, fourth column) many subjects chose the same plot, that had a higher $p$-value. Some of these anomalous lineups were selected for follow-up studies using an eye-tracker to investigate what people were looking at.

@@

Visual inference is like other types of computationally intensive test procedures, such as permutation tests, where the statistics from the data are compared with those from null data. But unlike these approaches visual inference is constrained to make the comparison with just a few draws $(m-1)$ from the null distribution. How this small set of null plots influences the observers choice is important for understanding the reliability of visual inference. If the actual data plot is very different from the null plots, then the null plots should not have much influence on the choice. Measuring the difference between the plots is almost impossible! But in the controlled setting here we can again use $p$-value as a proxy for similarity and difference.

In the previous section we noticed that observers tend to pick the plot with the lowest $p$-value. If a null plot has data that, by chance because the sample is a random draw from the null, has a smaller $p$-value than the actual data, we would expect that there is a strong possibility that an observer will pick this null plot instead of the actual data plot. Figure \ref{fig:pval_difference} investigates this for the data collected in experiments 1 and 2. The plots display the proportion of observers who choose the actual data plot against the $p$-value of the actual data minus the lowest seen in the null data. A small negative value of the difference means that the $p$-value of the actual data is smaller than that of the null data. If there is a big negative difference we'd expect to see that the observers would more often make the correct choice of plot. When there is a large positive difference, observers fail to pick the correct plot. If however, there is little difference between the $p$-values then we'd expect to see more confusion in the choices.

@@

 \green{Is this really pct correct, or 1-pct correct? \blue{This is proportion correct}. When there is a big diff in p-value almost no-one gets it right! Maybe it is just a scale issue?} \blue{In fact this is what we expect. If there is large difference that means data plot $p$-value is large while minimum $p$-value is very small. This should lead people to pick the wrong plot. Our minimum $p$-value assumption basically says this fact, right?}. We see that when there is a big difference between the $p$-values observers choose the actual data plot more often.

@@

A little known error amongst statisticians is what was coined as Type III error in \citet{mosteller:48}. This occurs when the null hypothesis is rejected for thw rong reason. Experiment 3 is prone to this type of error. Observers were asked to identify the plot that had the largest slope, either positive or negative. But the actual data plot had a cluster of points, the contamination that forced the conventional test to fail to see any trend. For the human eye this cluster of points is probably as visible as the association between the remaining points, leading the observer to detect the actual data plot by looking for the cluster instead of the slope. This would be considered to be a Type III error because it is not what was requested, but it still gives the correct answer. This is not a problem with the lineup protocol, because we want observers to react to anything that sets the the actual data plot apart. Here, in the controlled but simple setting, it is of interest to see what observers cued to in the actual data plot.

Figure \ref{fig:power_loess_effect} indicates that Type III error might be occurring in experiment 3: correct identification of the actual data plot is not positively associated with effect size. Teasing this out of the results is possible by looking at the reasons observers gave for their choices. Observers were provided with four possible reasons to use for their choice:

@@

\noindent with the possibility to use more than one. The task requested subjects to identify the plot that had the largest slope, which would correspond to choosing ``visible trend'' (2) as the reason for their choice. Reasons 1 or 3 would be indicative of Type III error. Figure \ref{fig:choice_reason} explores the reasons subjects gave for their choices. If there were no Type III errors committed, we would expect that people overwhelmingly using ``visible trend'' as their reason, or at least, when they use this reason they overwhelmingly correctly choose the actual data plot. This is not what we see. At left, are the reasons subjects gave for their choices --- 123 means that they gave all three reasons. The horizontal axis shows proportion of times that subjects correctly chose the actual data plot, and the reasons are sorted from most accurate to least accurate. The size of the point corresponds to the number of subjects putting this as the reaosn. Subjects that chose all three reasons almost always chose the actual data plot. This was followed by using 1 and 3, and then 1 and 4. The most common reasons given were reasons 1-3 individually, and the accuracy for these reasons ranged from 75\% for reason ``most different plot'' to 60\% for ``visible trend''.
At right is a simplified view, containing just the four possible reasons -- if the subject chose one of these, regardless if they also chose another reason it is counted. ``Visible trend'' comes in third. This is strong evidence that for many subjects even though they are correctly choosing the data plot, often they are cueing to other structure in the plot than the trend, making a Type III error.

%It is interesting to calculate the proportion of observers who choose the correct plot, but use reasons 1 or 3, relative to reason 2. Figure \ref{fig:choice_reason} examines this. For observers who chose the actual data plot, the proportions for the three reasons are shown. It can be seen that the most common reason was ``most different plot'' (1), followed by clustering visible (3), with ``visible trend'' coming in third. \green{What about other??} \blue{Not many people chose other as we see in Figure \ref{fig:choice_reason}.}

@@

@@

For visual inference, making a Type III, is not actually a problem. It is only a possibility in this ecperiment because we are working with known structure. In the real setting, we are excited to see observers detecting the actual data plot, and curious about how they detect it, with all possible reasons encapsulated in the alternative hypothesis. This does show the importance of the qualitative answers that observers give for their choices.

@@

In practice there is no correct answer. In these experiments we controlled the effets that observers saw, so there was a correct choice in each case. In the real setting, there is not correct choice, just whether the actual data plot stands apart from the rest, so that the observer can find it. And observers would provide open-ended answers for their choice of plot.

>>>>

After many small pilot studies with local personnel, it was clear that some care was needed to set up the human subjects experiments. It was best for a observer or a subject to see a block of 10 lineups with varying difficulty, with a reasonable number of ``easy'' lineups. The explanations about each experiment (below) includes an explanation of how the lineups were sampled and provided to the subjects, for each of the three experiments.

@@

The experiment is designed to study the ability of human subjects to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated a range of values of $\beta_2~ (=0, 1, 3, 5, 7, 8, 10, 16)$, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical conventional test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making a side-by-side boxplot of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and these were plotted in the same way as side-by-side boxplots. The actual data plot was randomly placed among these null data plots to produce the lineup. Figure \ref{fig:test_category} is an example of one of these lineups. The actual data plot is number 11 which was generated for $n$=300, $\beta_2$=15 and $\sigma$=12.

@@

%With multiple subjects it is possible to estimate the visual $p$-value, and a measure of plot signal strength. The visual $p$-value is calculated from the binomial formula (Equation \ref{binom}). Assuming $H_o$ is true, the probability of picking the actual data plot by chance is $1/m$, and the visual $p$-value is the probability that $x$ or more subjects select that plot. Figure \ref{fig:pval_definition} displays this visual $p$-value for each lineup against the conventional $p$value.

%Similary, the strength of the signal, is calculated using the proportion of subjects picking actual data plot. It is closely related to the power of the test. From Equation \ref{eqn:power_estimate}  we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives the signal strength estimate as  
%\[
%\hat p_D= (1- {u}/{K})/({m-1})
%\]
%where $K$ is the total number of independent evaluations on a lineup and $u$ is the number of successful evaluations. Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with conventional test $p$-values, on a square root scale. For experiment 3, technically, there are no conventional $p$-values, but the $p$-values are computed in the same manner as experiment 2, assuming that the data satifies the assumptions. The blue line is a loess fit to the data, and the grey line indicates equality of the two axes. The relationship between the conventional $p$-value and the plot signal strength is very strong, particularly when the values are small. As the conventional $p$-value increases, the estimated signal strength increases -- subjects find it harder to detect the actual data plot at larger $p$-values. The rise is very sharp for experiment 1, which indicates that subjects have a harder time detecting the actual data plot for even small $p$-values.

@@

One assumption made in order to calculate the effect of lineup size in the calculations of visual $p$-value and signal strength was that subjects would tend to pick the plot in the lineup that had the strongest signal. In experiments 1 and 2, this corresponds to the plot with the smallest $p$-value. This can be examined for the data collected, to see if this assumption is, indeed, reasonable. %In Section \ref{sec:size} in order to make comparisons with conventional tests an assumption was made that the subject would pick the plot in the lineup that had the lowest $p$-value. In this section we take a look at the picks that subjects made, to see if this is, indeed, true.

Figures \ref{fig:P-val_log} and \ref{fig:P-val_log2} show the selection counts and $p$-values of all plots for each lineup in experiments 1 and 2, respectively. In each of these plots, the log of $p$-value for each plot in the lineup is plotted horizontally and the count of the number of subjects who selected this plot is displayed vertically. The dot represents this point, and the lines connect vertically with 0 count. Each row corresponds to a different $\beta$ value ordered top to bottom from lowest to highest. Columns correspond to the other levels of the experiment, sample size $(n)$, error standard deviation $(\sigma)$ and replicate. (Empty cells indicate no lineup shown with this combination.) Red indicates the plot with the lowest $p$-value in the lineup. In both experiments as $\beta$ gets larger (stronger effect) people tended to select the plot with the lowest $p$-value. This is less so the case when $\beta$ is small, when standard deviation is large and sample size is small. The results are clearer for experiment 2, that used a continuous covariate. But even when they didn't pick the plot with the lowest $p$-value they tended to oscillate their choices between the several low $p$-value plots. So for most subjects, the assumption that they pick the plot with the smallest $p$-value would appear to be reasonable, and the actual power of the visual test should be close to the expected power.

@@

There are some noticeable exceptions to this rule. In experiment 1, when $\beta=5, n=300, sd=12, rep=3$, people tended to pick the plot with the second smallest $p$-value.  Same experiment, when $\beta=0, n=100, sd=5, rep=1$ people overwhelming chose a plot with much larger $p$-value. For several of these exceptions, along with several easy lineups, we followed up with a new experiment using an eye-tracker to examine what the subjects were looking at in making their choices. \cite{*** yifan's CC needs to be made into a tech report, and referenced here.}

Experiment 3 does not have $p$-values associated with each plot, so the assumption can only be tested with the first two experiments. Practically, this is not an assumption that is explicitly made, because there will not be a $p$-value available. Rather it is assumed that the subject has the ability to select the plot with the most extreme structure. A few of the lineups indicated that occasionally people cue on different patterns, for example, in experiment 1 the lineup $\beta=0, n=100, sd=5, rep=1$ (first row, fourth column) many subjects chose the same plot, that had a higher $p$-value. Some of these anomalous lineups were selected for follow-up studies using an eye-tracker to investigate what people were looking at.

@@

Visual inference is like other types of computationally intensive test procedures, such as permutation tests, where the statistics from the data are compared with those from null data. But unlike these approaches visual inference is constrained to make the comparison with just a few draws $(m-1)$ from the null distribution. How this small set of null plots influences the subjects choice is important for understanding the reliability of visual inference. If the actual data plot is very different from the null plots, then the null plots should not have much influence on the choice. Measuring the difference between the plots is almost impossible! But in the controlled setting here we can again use $p$-value as a proxy for similarity and difference.

In the previous section we noticed that subjects tend to pick the plot with the lowest $p$-value. If a null plot has data that, by chance because the sample is a random draw from the null, has a smaller $p$-value than the actual data, we would expect that there is a strong possibility that an subject will pick this null plot instead of the actual data plot. Figure \ref{fig:pval_difference} investigates this for the data collected in experiments 1 and 2. The plots display the proportion of subjects who choose the actual data plot against the $p$-value of the actual data minus the lowest seen in the null data. A small negative value of the difference means that the $p$-value of the actual data is smaller than that of the null data. If there is a big negative difference we'd expect to see that the subjects would more often make the correct choice of plot. When there is a large positive difference, subjects fail to pick the correct plot. If however, there is little difference between the $p$-values then we'd expect to see more confusion in the choices.

@@

 \green{Is this really pct correct, or 1-pct correct? \blue{This is proportion correct}. When there is a big diff in p-value almost no-one gets it right! Maybe it is just a scale issue?} \blue{In fact this is what we expect. If there is large difference that means data plot $p$-value is large while minimum $p$-value is very small. This should lead people to pick the wrong plot. Our minimum $p$-value assumption basically says this fact, right?}. We see that when there is a big difference between the $p$-values subjects choose the actual data plot more often.

@@

A little known error amongst statisticians is what was coined as Type III in \citet{mosteller:48}. This occurs when the null hypothesis is rejected for thw rong reason. Experiment 3 is prone to this type of error. Subjects were asked to identify the plot that had the largest slope, either positive or negative. But the actual data plot had a cluster of points, the contamination that forced the conventional test to fail to see the association. For the human eye this cluster of points is probably as visible as the association between the remaining points, leading the subject to detect the actual data plot by looking for the cluster instead of the slope. This would be considered to be a Type III error because it is not what was requested, but it still gives the correct answer. This is not a problem with the lineup protocol, because we want subjects to react to anything that sets the the actual data plot apart. Here, in the controlled but simple setting, it is of interest to see what subjects cued to in the actual data plot.

Figure \ref{fig:power_loess_effect} indicates that Type III error might be occurring in experiment 3: correct identification of the actual data plot is not positively associated with effect size. Teasing this out of the results is possible by looking at the reasons subjects gave for their choices. Subjects were provided with four possible choices

@@

\noindent It was possible to choose more than one reason. The question begged for ``visible trend'' (2) to be the reason that subjects gave for correctly choosing the actual data plot. Either reasons 1 or 3 would be indicative of Type III error. It is interesting to calculate the proportion of subjects who choose the correct plot, but use reasons 1 or 3, relative to reason 2. Figure \ref{fig:choice_reason} examines this. For subjects who chose the actual data plot, the proportions for the three reasons are shown. It can be seen that the most common reason was ``most different plot'' (1), followed by clustering visible (3), with ``visible trend'' coming in third. \green{What about other??} \blue{Not many people chose other as we see in Figure \ref{fig:choice_reason}.}

@@

\green{I find this plot really confusing! Can we just look at whether they selected choice ``clustering visible'' or not, if they were correct?} \blue{Is the right hand side plot in Figure \ref{fig:choice_reason} ok? Should we remove the left plot?}

@@

@@

In practice there is no correct answer. In these experiments we controlled the effects that observers saw, so there was a correct choice in each case. In the real setting, there is not correct choice, just whether the actual data plot stands apart from the rest, so that the observer can find it. And subjects would provide open-ended answers for their choice of plot.
